A data-driven, pixel-level model to improve the precision of \Kepler\ photometry

### 2013-10-15 documents/plm.tex
todos, tweak

The photometric measurements of stars made by the \Kepler\ Satellite are precise enough
  to permit discovery of exoplanet transits with depths smaller than $10^{-4}$.
This precision results from great spacecraft stability,
  supplemented by various methods for removing small residual spacecraft-induced and stellar-variability trends in the brightnesses,
  either filtering the data (with things like median filters; CITE)
  or fitting the data with flexible models (like polynomials or splines or Gaussian Processes; CITE).
When employed in the service of exoplanet search and characterization,
  these methods are usually agnostic about whether photometric variations originate in the spacecraft or in the star itself;
  that is, they obliterate intrinsic stellar variability along with spacecraft issues.

In general, there are many reasons for apparent photometric variability in a \Kepler\ source.
There is intrinsic stellar variability,
  which is of interest to some and a nuisance to others.
There is also variability of overlapping fainter stars;
  that is, confusion noise combined with variability of the confusing sources.
There are small changes in spacecraft pointing,
  which leads to slightly different illumination of the focal-plane pixels,
  and thus different sensitivity to errors (problems) in the device flat-field or sensitivity map.
There are also \emph{intra-pixel} sensitivity variations that can contribute (CITE WHITEPAPER, SPITZER).
There are small changes in spacecraft temperature,
  which lead to point-spread function (PSF) and differential (across the focal plane) pointing changes.
These also lead to changes in pixel and intra-pixel illumination.
Stellar proper motion, geometric parallaxes, and differential stellar aberration as the spacecraft orbits all do more of the same.
There is electronic cross-talk between detectors and charge-transfer inefficiency;
  these can effectively transfer variability from one source to another.
There are additional electronics effects like ``rolling bands'' that put additional features into lightcurves.
There are also probably changes to the detector sensitivity with temperature and time,
  possibly illimunation-history effects,
  and possibly sources of variability not yet considered.
The remarkable thing about \Kepler\ is that it is trying to measure stars at a level of precision
  much higher than ever previously attempted;
  new effects really \emph{must} appear at some point.
In \figurename~\ref{fig:pixelpatch}, we show the pixel-level variability in the \Kepler\ data
  near one bright star that shows minimal intrinsic variability;
  this figure highlights the spacecraft-induced effects.

@@

One such method---one that is designed to model spacecraft-induced problems
  but \emph{not} interfere with measurements of stellar variability---%
  is the \Kepler\ Presearch (\sic) Data Conditioning (PDC) method (CITE).
The PDC ``de-trends'' the \Kepler\ lightcurves by fitting them with a small set of basis lightcurves
  generated from a principal components analysis (PCA) of filtered lightcurves.
That is, it uses data to model data,
  regularizing the fit (and avoiding over-fitting) by filtering and restricting the dimensionality (through PCA).
The method proposed here, \name, is very similar in spirit to the PDC.
The main differences are
  (1)~that \name\ works in the pixel domain, not the lightcurve domain, so it has access to more fine-grained information,
  (2)~that \name\ has far more freedom (far more parameters) than the PDC
  but it strictly avoids over-fitting the lightcurves on exoplanet-transit time-scales
  through strong regularization and a train-and-test framework, and
  (3)~that when \name\ fits a star, there is no possibility of any contribution of the star itself to the fit basis vectors.
The two methods (\name\ and PDC) are similar however,
  in that they both make the assumption that whatever spacecraft effects are imprinting variability on a stellar lightcurve
  must also be imprinting similar or related variability on other lightcurves.
They also assume that the relationships can be captured with linear models.
By going to the pixel level (unlike the PDC, which works at the photometry level),
  \name\ makes it easier for the model to capture variability
  that is coming through variations in the centroids and point-spread function
  from spacecraft pointing, roll, and temperature.

Before we start, a few reminders about the \Kepler\ data are in order:
The satellite always observes precisely the same field, at fixed pointing (as closely as possible).
The satellite is rolled by 90~deg every 90-ish days (for Solar Angle constraints).
The PSF varies strongly across the field and is badly sampled.
The stars span a huge range in brightness;
  some of \Kepler's most important targets even saturate the device and bleed charge.
The stellar photometry returned by the \Kepler\ SAP and PDC pipelines is based on
  straight, \emph{unweighted} sums of pixels in small patches centered on the stellar centroid.
We will return to this latter point at the end;
  this kind of photometry cannot be optimal;
  it must be possible to do a better job with the photometric measurements.
That's beyond the scope of this paper but a place for a valuable intervention on the \Kepler\ data.

>>>>

The photometric measurements of stars made by the \Kepler\ Satellite are precise enough
  to permit discovery of exoplanet transits with depths smaller than $10^{-4}$.
This precision results from great spacecraft stability,
  supplemented by various methods for removing small residual spacecraft-induced and stellar-variability trends in the brightnesses,
  either filtering the data (with things like median filters; \todo{citations})
  or fitting the data with flexible models (like polynomials or splines or Gaussian Processes; \todo{citations}).
When employed in the service of exoplanet search and characterization,
  these methods are usually agnostic about whether photometric variations originate in the spacecraft or in the star itself;
  that is, they obliterate intrinsic stellar variability along with spacecraft issues.

In general, there are many reasons for apparent photometric variability in a \Kepler\ source.
There is intrinsic stellar variability,
  which is of interest to some and a nuisance to others.
There is also variability of overlapping fainter stars;
  that is, confusion noise combined with variability of the confusing sources.
There are small changes in spacecraft pointing,
  which leads to slightly different illumination of the focal-plane pixels,
  and thus different sensitivity to errors (problems) in the device flat-field or sensitivity map.
There are also \emph{intra-pixel} sensitivity variations that can contribute (\todo{cite white paper; Spitzer results}).
There are small changes in spacecraft temperature,
  which lead to point-spread function (PSF) and differential (across the focal plane) pointing changes.
These also lead to changes in pixel and intra-pixel illumination.
Stellar proper motion, geometric parallaxes, and differential stellar aberration as the spacecraft orbits all do more of the same.
There is electronic cross-talk between detectors and charge-transfer inefficiency;
  these can effectively transfer variability from one source to another.
There are additional electronics effects like ``rolling bands'' that put additional features into lightcurves.
There are also probably changes to the detector sensitivity with temperature and time,
  possibly illimunation-history effects,
  and possibly sources of variability not yet considered.
The remarkable thing about \Kepler\ is that it is trying to measure stars at a level of precision
  much higher than ever previously attempted;
  new effects really \emph{must} appear at some point.
In \figurename~\ref{fig:pixelpatch}, we show the pixel-level variability in the \Kepler\ data
  near one bright star that shows minimal intrinsic variability;
  this figure highlights the spacecraft-induced effects.

@@

One such method---one that is designed to model spacecraft-induced problems
  but \emph{not} interfere with measurements of stellar variability---%
  is the \Kepler\ Presearch (\sic) Data Conditioning (PDC) method (\todo{citation}).
The PDC ``de-trends'' the \Kepler\ lightcurves by fitting them with a small set of basis lightcurves
  generated from a principal components analysis (PCA) of filtered lightcurves.
That is, it uses data to model data,
  regularizing the fit (and avoiding over-fitting) by filtering and restricting the dimensionality (through PCA).
The method proposed here, \name, is very similar in spirit to the PDC.
The main differences are
  (1)~that \name\ works in the pixel domain, not the lightcurve domain, so it has access to more fine-grained information,
  (2)~that \name\ has far more freedom (far more parameters) than the PDC
  but it strictly avoids over-fitting the lightcurves on exoplanet-transit time-scales
  through strong regularization and a train-and-test framework, and
  (3)~that when \name\ fits a star, there is no possibility of any contribution of the star itself to the fit basis vectors.
The two methods (\name\ and PDC) are similar however,
  in that they both make the assumption that whatever spacecraft effects are imprinting variability on a stellar lightcurve
  must also be imprinting similar or related variability on other lightcurves.
They also assume that the relationships can be captured with linear models.
By going to the pixel level (unlike the PDC, which works at the photometry level),
  \name\ makes it easier for the model to capture variability
  that is coming through variations in the centroids and point-spread function
  from spacecraft pointing, roll, and temperature.

Before we start, a few reminders about the \Kepler\ data are in order:
The satellite always observes precisely the same field, at fixed pointing (as closely as possible).
The satellite is rolled by 90~deg every 90-ish days (to satisfy Solar-Angle constraints).
The PSF varies strongly across the field and is badly sampled.
The stars span a huge range in brightness;
  some of \Kepler's most important targets even saturate the device and bleed charge.
The stellar photometry returned by the \Kepler\ SAP and PDC pipelines is based on
  straight, \emph{unweighted} sums of pixels in small patches centered on the stellar centroid.
We will return to this latter point at the end;
  this kind of photometry cannot be optimal;
  it must be possible to do a better job with the photometric measurements.
That's beyond the scope of this paper but a place for a valuable intervention on the \Kepler\ data.
